
---
title: "Detecting batch effect - report 1"
author: "Qunhua Li"
date: "6/5/2019"
output: html_document
---

Batch effect can be a cause of lack of reproducibility. Here we demonstrate batch effect with the data set in [Spielman et al. (2007)](http://www.ncbi.nlm.nih.gov/pubmed/17206142)[^1].  

```{r message=FALSE, warning=FALSE}
# Run this part if you need install the following packages
# If R version is lower than 3.5, use the next 3 lines
# source("http://bioconductor.org/biocLite.R") 
# BiocInstaller::biocLite("Biobase") 
# BiocInstaller::biocLite("genefilter")

# If R version is 3.5+, use the following three lines
# if (!requireNamespace("BiocManager", quietly = TRUE))
# install.packages("BiocManager")
# BiocManager::install(c("Biobase", "genefilter"))

# run this line regardless the R version 
# devtools::install_github("genomicsclass/GSE5859")

library(Biobase)
library(rafalib)
library(GSE5859)
library(genefilter)

data(GSE5859)
```

```{r }
# gene expression levels  
geneExpression = exprs(e)
```

# Data collection dates
```{r }
# print date info
sampleInfo = pData(e)
head(sampleInfo$date)

# number of data collection dates
length(unique(sampleInfo$date))
```

The data were collected in `r length(unique(sampleInfo$date))` different dates.

Check years that the data from different ethnic groups were collected. We can immediately see that year and ethnicity are almost completely confounded.
```{r }
# Print a table of year vs ethnic groups
year = factor( format(sampleInfo$date,"%y") )
tab = table(year,sampleInfo$ethnicity)
print(tab)
```


# Differential expression analysis
```{r }
##remove control genes
out <- grep("AFFX",rownames(geneExpression))
```

- We first use a t-test to perform differential expression analysis between CEU and ASN.
```{r }
eth <- sampleInfo$ethnicity
ind<- which(eth%in%c("CEU","ASN"))
res1 <- rowttests(geneExpression[-out,ind],droplevels(eth[ind]))
```

- We then perform the same analysis between 2002 vs 2003 within CEU.
```{r }
ind <- which(year%in%c("02","03") & eth=="CEU")
res2 <- rowttests(geneExpression[-out,ind],droplevels(year[ind]))
```

- We plot histograms for p-values as in Akey et al (2007)[^2]. 
```{r }   
mypar(1, 2)
hist(res1$p, main="p-value of CEU vs ASN")
hist(res2$p, main="p-value of 2002 vs 2003 for CEU")
```

There are a lot of differentially expressed genes between the CEU samples collected in 2002 and 2003. 

[^1]: Spielman, R. et al. Common genetic variants account for differences in gene expression among ethnic groups. *Nature Genetics*. **39**, pages 226–231 (2007)
[^2]: Akey, J. et al. On the design and analysis of gene expression studies in human populations. *Nature Genetics*. **39**, pages 807–808 (2007)